Redundant calibration as a complex optimization problem

### 2017-01-09 paper.tex
newest version

\subsection{Alternating Direction Implicit Method}
\label{sec:adi}

%In Section~\ref{sec:adi} we show how the ADI method used for skymodel-based calibration relates to the redundant ADI method presented in \citet{Marthi2014}. 
In this section we re-derive the ADI method presented in \citet{Marthi2014}. In \citet{Marthi2014} the ADI method is derived by taking the derivative of the 
objective function $\Lambda$ relative to $\bg$ and $\by$ and setting the result to naught. When derived in this way, its relation to the recent works \citep{Salvini2014,Smirnov2015} are unclear.
We will use the diagonal-approximate approach proposed in \citet{Smirnov2015} to show that the ADI algorithm presented in \citet{Marthi2014} is in actual fact redundant 
\textsc{StEfCal} \citep{Salvini2014}. Perusing through Appendix~\ref{sec:sbc} is advised before continuing this section.

@@

Optimality results of ADI were presented in \citet{Marthi2014}. We do not repeat their analysis here. We however, do present our own ADI results in Fig.~\ref{fig:prec_error} that confirm that the
ADI method can accurately estimate the antenna gains and the observed visibilities at different SNR values. To quantity the exact error in our estimates we used the precentage error $\beta$ defined 
as 
\begin{equation}
\label{eq:beta}
\beta = \frac{\sum_{pq} |v_{pq} - \widehat{v}_{pq} |^2}{v_{pq}}.
\end{equation}
We have already defined $v_{pq}$ in Eq.~\eqref{eq:vec_definitions}. Moreover, in Eq.~\eqref{eq:beta}, $\widehat{v}_{pq}$ denotes the ADI estimate of $v_pq$.

@@

\subsection{Preconditioned Conjugate Gradient Method}
\label{sec:pcg}
In this section we investigate the computational complexity of the PCG method when applied to the redundant calibration problem. 
The PCG method was presented in Section~\ref{sec:conj_grad}.
The PCG method differs from the ADI method presented in Section~\ref{sec:adi} in that in the case of PCG we compute the exact inverse of the modified Hessian $\bmH$ instead 
of using some sort of approximation as is the case for ADI. As we mentioned in Section~\ref{sec:pcg} there
are two factors which influence the computational complexity of the PCG method, namely the spectral condition number and the sparsity of the matrix we wish to invert (see Eq.~\eqref{eq:cg_bound}). 
We used a $\lambda$ value of two, implying we used Eq.~\eqref{eq:LM_update} instead of Eq.~\eqref{eq:GN_update} throughout this section.
We respectively discuss the spectral condition number and the sparsity of the modified Hessian $\bmH$, which is a non-singular matrix, in Section~\ref{sec:scn} and Section~\ref{sec:sparsity}.
It is interesting to note, that the Hessian $\bH$ itself is singular which means that if we are to use Eq.~\eqref{eq:GN_update} instead we would need to be able to invert a singular matrix. In Section~\ref{sec:normal},
we showed that the PCG method can in fact be used to invert a singular Hessian $\bH$.

\subsubsection{Spectral Condition Number}
\label{sec:scn}
In Section~\ref{sec:precon} we mentioned that the convergence properties of CG can be dramatically improved through preconditioning. Inspecting Fig.~\ref{fig:hessian} reveals that the Jacobian preconditioner
is a good candidate for the redundant use case. In Fig.~\ref{fig:kappa} we see that this choice of preconditioner performs very well, the value of the spectral condition number $\kappa$
becomes constant and independent of problem size, i.e. the number of antennas in the array. Preconditioning effectively eliminates it from Eq.~\eqref{eq:cg_bound}. This result
is confirmed by Fig.~\ref{fig:cg_itr}.

%kappa + itr plots
\begin{figure*}
\centering
\subfigure[$\kappa$]{\includegraphics[width=0.47\textwidth]{./kappa.pdf}\label{fig:kappa}}
% V_R_3.pdf: 585x441 pixel, 72dpi, 20.64x15.56 cm, bb=0 0 585 441
\subfigure[Iterations required before and after Preconditioning]{\includegraphics[width=0.47\textwidth]{./cg_itr.pdf}\label{fig:cg_itr}}
\caption{The graphs presented here were generated using the second experimental setup in Table~\ref{tab:ch_parm}. In the left panel we have plotted the spectral condition number $\kappa$ of the modified Hessian $\bmH$, before (magenta and red curves) and after preconditioning (blue curve), at different SNR values as a function of the
number of antennas $N$ in the array. This plot shows us that we can drastically reduce the condition number of $\bmH$ by preconditioning. Moreover, $\kappa$ is also
independent of the problem size (the number of antennas in the array). The fact that $\kappa$ is small and independent of problem size (effectively constant) implies that it simply drops out 
of Eq.~\eqref{eq:cg_bound}. The right hand panel confirms this result. In the right hand panel we have plotted the number of major iterations (line 6 of Algorithm \ref{algo:CG}) required by the conjugate 
gradient method to invert $\bmH$, before (magenta and red curves) and after preconditioning (blue and green curves), at different SNR values as a function of the number of antennas $N$ in the array.  
The right hand panel therefore shows us that the number of major iterations needed by the preconditioned conjugate gradient method to invert $\bmH$ is independent of the number of antennas in the array.
The number of major iterations are also much smaller than the dimension of $\bmH$. 
\label{fig:kappa_itr}} 
\end{figure*}

@@

The second factor that determines the computational complexity of the conjugate gradient method when used to perform redundant calibration is the amount of non-zero
entries in the modiefied Hessian matrix $\bmH$. 
%We have also shown in section that the first factor, the spectral condition number of $\bH$, can be reduced 
%to a constant by employing preconditioning and can therefore be completely eliminated from the computational complexity equation. 
A useful matrix quantity which we will use in this section is the sparsity factor $\gamma$ of a matrix, which is the ratio between the amount of zero entries and the total amount of entries in a matrix. We 
can therefore define $\gamma$ as follows:
\begin{equation}
 \gamma = \left (1 - \frac{m}{P^2} \right ) 
\end{equation}
It is trivial to determine an analytic expression for the sparsity ratio of $\bmH$ when our redundant array is in a regular east-west geometric configuration
The sparsity and the asymptotic sparsity ratio of $\bmH$ are equal to
\begin{align}
\gamma &= \frac{5N^2-7N+3}{8N^2-8N+2} & \gamma_{\infty} &= \lim_{N\rightarrow \infty}\gamma = \frac{5}{8} \label{eq:gamma}. 
\end{align}
However when our redundant array makes use of a more complicated geometric layout it becomes much more difficult to construct such expressions as $\bmH$ itself 
becomes much harder to express analytically.

In Fig.~\ref{fig:gamma} we plot the sparsity ratio for three different geometric layouts as a function of the number antennas in the array. Although, we can not explicitly calculate 
the limit of the hexagonal and square layouts, Fig.~\ref{fig:gamma} seems to indicate that these limits do exist.

We can now use the sparsity factor to compute the order of the computational complexity of inverting $\bmH$ with
\begin{equation}
P^{c} = (1 - \gamma)P^2.
\end{equation}
Solving for $c$ we obtain
\begin{align}
c &= \log_{P}(1 - \gamma) + 2 & c_{\infty} &= \lim_{N\rightarrow \infty} c = 2. \label{eq:c}
\end{align}
In Fig.~\ref{fig:c} we plot the value of $c$ for three different geometric layouts as a function of the number antennas in the array.
It is clear from Eq.~\eqref{eq:c} and Fig.~\ref{fig:c} that the computational complexity, under the assumption that $\lim_{N\rightarrow \infty} \gamma$ exists, is asymptotically bounded by $O(P^2)$. However, the computational complexity converges 
very slowly to its asymptotic value, and in general is equal to $O(P^{c})$, where $c \sim 1.7$ if a Hexagonal geometric layout containing less than 200 antennas is employed.

\begin{figure*}
\centering
\subfigure[$\gamma$]{\includegraphics[width=0.47\textwidth]{./sparsity.pdf}\label{fig:gamma}}
% V_R_3.pdf: 585x441 pixel, 72dpi, 20.64x15.56 cm, bb=0 0 585 441
\subfigure[$c$]{\includegraphics[width=0.46\textwidth]{./comp_order.pdf}\label{fig:c}}
\caption{In the left panel we plot the sparsity ratio $\gamma$ (the number of zero entries relative to the total number of entries in a matrix) of the modified Hessian $\bmH$ as a function of the number 
of antennas $N$ in the array for different geometries, namely an hexagonal grid (blue circles), a square grid (green crosses) and a regular east-west grid (red circles). For the regular east-west grid we are able to construct an analytical expression for $\gamma$ (the dashed red line). Taking the limit 
of this expression, Eq.~\eqref{eq:gamma}, as $N\rightarrow \infty$ produces the black dotted line. In the right panel we plot the order of the computational cost $c$ for inverting $\bmH$ as a function of the number 
of antennas in the array for different geometries. We use the same color scheme in the right and left panels. We compute $c$ using Eq.~\eqref{eq:c}. The red dashed line 
in the right panel is associated with the theoretical curve of $c$ we computed for the east-west regular grid. The limit of the red-dashed curve is two, which we 
have plotted with a black dotted line.
\label{fig:sparsity}} 
\end{figure*}

>>>>

\subsection{Alternating Direction Implicit Method}
\label{sec:adi}
%In Section~\ref{sec:adi} we show how the ADI method used for skymodel-based calibration relates to the redundant ADI method presented in \citet{Marthi2014}. 
In this section we re-derive the ADI method presented in \citet{Marthi2014}. In \citet{Marthi2014} the ADI method is derived by taking the derivative of the 
objective function $\Lambda$ relative to $\bg$ and $\by$ and setting the result to naught. When derived in this way, its relation to the recent works \citep{Salvini2014,Smirnov2015} are unclear.
We will use the diagonal-approximate approach proposed in \citet{Smirnov2015} to show that the ADI algorithm presented in \citet{Marthi2014} is in actual fact redundant 
\textsc{StEfCal} \citep{Salvini2014}. Perusing through Appendix~\ref{sec:sbc} is advised before continuing this section.

@@

Optimality results of ADI were presented in \citet{Marthi2014}. We do not repeat their analysis here. We however, do present our own ADI results in Fig.~\ref{fig:prec_error} that confirm that the
ADI method can accurately estimate the antenna gains and the observed visibilities at different SNR values. To quantity the exact error in our estimates we used the precentage error $\beta$ defined 
as 
\begin{equation}
\label{eq:beta}
\beta = \frac{\sum_{pq} |v_{pq} - \widehat{v}_{pq} |^2}{\sum_{pq} |v_{pq}|^2}.
\end{equation}
We have already defined $v_{pq}$ in Eq.~\eqref{eq:vec_definitions}. Moreover, in Eq.~\eqref{eq:beta}, $\widehat{v}_{pq}$ denotes the ADI estimate of $v_{pq}$.

@@

\subsection{Preconditioned Conjugate Gradient Method}
\label{sec:pcg}
In this section we investigate the computational complexity of the PCG method when applied to the redundant calibration problem. 
As we mentioned in Section~\ref{sec:pcg} there are two factors which influence the computational complexity of the PCG method, namely the spectral condition number and the sparsity of the matrix we wish to invert (see Eq.~\eqref{eq:cg_bound}). 
We used a $\lambda$ value of two, implying we used Eq.~\eqref{eq:LM_update} instead of Eq.~\eqref{eq:GN_update} throughout this section.
We respectively discuss the spectral condition number and the sparsity of the modified Hessian $\bmH$, which is a non-singular matrix, in Section~\ref{sec:scn} and Section~\ref{sec:sparsity}.
It is interesting to note, that the Hessian $\bH$ itself is singular which means that if we are to use Eq.~\eqref{eq:GN_update} instead we would need to be able to invert a singular matrix. In Section~\ref{sec:normal},
we showed that the PCG method would in fact be able to handle such a scenario.

\subsubsection{Spectral Condition Number}
\label{sec:scn}
In Section~\ref{sec:precon} we mentioned that the convergence properties of CG can be dramatically improved through preconditioning. Inspecting Fig.~\ref{fig:hessian} reveals that the Jacobian preconditioner
is a good candidate for the redundant use case. In Fig.~\ref{fig:kappa} we see that this choice of preconditioner performs very well, the value of the spectral condition number $\kappa$
becomes constant and independent of problem size, i.e. the number of antennas in the array. Preconditioning therefore effectively eliminates $\kappa$ from Eq.~\eqref{eq:cg_bound}. This result
is confirmed by Fig.~\ref{fig:cg_itr}. Fig.~\ref{fig:cg_itr} shows us that the number of major iterations needed by the preconditioned conjugate gradient method to invert $\bmH$ is independent of the number of antennas in the array.
Moreover, the number of major iterations required are also much smaller than the dimension of $\bmH$.

%kappa + itr plots
\begin{figure*}
\centering
\subfigure[$\kappa$]{\includegraphics[width=0.47\textwidth]{./kappa.pdf}\label{fig:kappa}}
% V_R_3.pdf: 585x441 pixel, 72dpi, 20.64x15.56 cm, bb=0 0 585 441
\subfigure[Iterations required before and after Preconditioning]{\includegraphics[width=0.47\textwidth]{./cg_itr.pdf}\label{fig:cg_itr}}
\caption{The graphs presented here were generated using the second experimental setup in Table~\ref{tab:ch_parm}. In the left panel we have plotted the spectral condition number $\kappa$ of the modified Hessian $\bmH$, before (magenta and red curves) and after preconditioning (blue curve), at different SNR values as a function of the
number of antennas $N$ in the array. The right hand panel confirms this result. In the right hand panel we have plotted the number of major iterations (line 6 of Algorithm \ref{algo:CG}) required by the conjugate 
gradient method to invert $\bmH$, before (magenta and red curves) and after preconditioning (blue and green curves), at different SNR values as a function of the number of antennas $N$ in the array.   
\label{fig:kappa_itr}} 
\end{figure*}

@@

The second factor that determines the computational complexity of the conjugate gradient method when used to perform redundant calibration is the amount of non-zero
entries in the modified Hessian matrix $\bmH$. 
%We have also shown in section that the first factor, the spectral condition number of $\bH$, can be reduced 
%to a constant by employing preconditioning and can therefore be completely eliminated from the computational complexity equation. 
A useful matrix quantity which we will use in this section is the sparsity factor $\gamma$ of a matrix, which is the ratio between the amount of zero entries and the total amount of entries in a matrix. We 
can therefore define $\gamma$ as follows:
\begin{equation}
 \gamma = \left (1 - \frac{m}{P^2} \right ) 
\end{equation}
It is trivial to determine an analytic expression for the sparsity ratio of $\bmH$ when our redundant array is in a regular east-west geometric configuration
The sparsity and the asymptotic sparsity ratio of $\bmH$ are equal to
\begin{align}
\gamma &= \frac{5N^2-7N+3}{8N^2-8N+2} & \gamma_{\infty} &= \lim_{N\rightarrow \infty}\gamma = \frac{5}{8} \label{eq:gamma}. 
\end{align}
However when our redundant array makes use of a more complicated geometric layout it becomes much more difficult to construct such expressions as $\bmH$ itself 
becomes much harder to express analytically.

In Fig.~\ref{fig:gamma} we plot the empirically determined sparsity ratios for three different geometric layouts as a function of the number antennas in the array. Although, we can not explicitly calculate 
the limit of the hexagonal and square layout curves, Fig.~\ref{fig:gamma} seems to indicate that they do exist. Moreover, we also plot the left most equation
of Eq.~\eqref{eq:gamma} (red dashed line) and the right most equation of Eq.~\eqref{eq:gamma} (black dotted curve) in Fig.~\ref{fig:gamma}.

We can now use the sparsity factor to compute the order of the computational complexity of inverting $\bmH$ (assuming the effect of $\kappa$ is negliable) with
\begin{equation}
P^{c} = (1 - \gamma)P^2.
\end{equation}
Solving for $c$ we obtain
\begin{align}
c &= \log_{P}(1 - \gamma) + 2 & c_{\infty} &= \lim_{N\rightarrow \infty} c = 2. \label{eq:c}
\end{align}
Note, that we can only compute $c_{\infty}$ in the rightmost equation if $\lim_{N\rightarrow \infty} \gamma$ exists. 
In Fig.~\ref{fig:c}, we plot the empiracally determined value of $c$ for three different geometric layouts as a function of the number antennas in the array.
Furhtermore, in Fig.~\ref{fig:c}, the analytic value of $c$ for the regular east-west grid is plotted as a dashed red curve, while the limit of this analytic curve is plotted with a dotted black curve.

Eq.~\eqref{eq:gamma}, Eq.~\eqref{eq:c} and Fig.~\ref{fig:sparsity} indicate that the computational complexity, under the assumption that $\lim_{N\rightarrow \infty} \gamma$ exists, is asymptotically bounded by $O(P^2)$.
However, the computational complexity converges very slowly to its asymptotic value, and in general is equal to $O(P^{c})$, where $c \sim 1.7$ if a Hexagonal geometric layout containing less than 200 antennas is employed.

\begin{figure*}
\centering
\subfigure[$\gamma$]{\includegraphics[width=0.47\textwidth]{./sparsity.pdf}\label{fig:gamma}}
% V_R_3.pdf: 585x441 pixel, 72dpi, 20.64x15.56 cm, bb=0 0 585 441
\subfigure[$c$]{\includegraphics[width=0.46\textwidth]{./comp_order.pdf}\label{fig:c}}
\caption{In the left panel we plot the sparsity ratio $\gamma$ (the number of zero entries relative to the total number of entries in a matrix) of the modified Hessian $\bmH$ as a function of the number 
of antennas $N$ in the array for different geometries, namely an hexagonal grid (blue circles), a square grid (green crosses) and a regular east-west grid (red circles). For the regular east-west grid we are able to construct an analytical expression for $\gamma$ (the dashed red line). Taking the limit 
of this analytic expression produces the black dotted line. In the right panel we plot the order of the computational cost $c$ for inverting $\bmH$ as a function of the number 
of antennas in the array for different geometries. We use the same color scheme in the right and left panels. The red dashed line 
in the right panel is associated with the theoretical curve of $c$ we computed for the east-west regular grid. The limit of the red-dashed curve is two, which we 
have plotted with a black dotted line.
\label{fig:sparsity}} 
\end{figure*}
